---
title: "DATA (IS) 605 FUNDAMENTALS OF COMPUTATIONAL MATHEMATICS - FALL 2016 - ASSIGNMENT 5"
author: "Robert Godbey"
date: "September 30, 2016"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: true
    theme: readable
    highlight: espresso
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Set 1

### (1) Work out some properties of the least squares solution

Consider the unsolvable system Ax = b as given below:

$\left[ \begin{array}{ccc} 1 & 0 \\ 1 & 1 \\ 1 & 3 \\ 1 & 4 \\ \end{array}\right]$
$\left[ \begin{array}{ccc} x_1 \\ x_2 \\ \end{array}\right] =$
$\left[ \begin{array}{ccc} 0 \\ 8 \\ 8 \\ 20 \\ \end{array}\right]$

#### a) Use R to compute $A^TA$ and $A^Tb$

```{r trans}
A <- matrix(c(1, 1, 1, 1, 0, 1, 3, 4), nrow=4)
b <- matrix(c(0, 8, 8, 20), nrow=4)
ata <- t(A) %*% A
ata
atb <- t(A) %*% b
atb
```


#### b) Solve for $\hat{x}$ in R using the above two computed matrices

Since $A^TA\hat{x} = A^Tb$ this give us 
$\left[ \begin{array}{ccc} 4 & 8 \\ 8 & 26 \\ \end{array}\right] \hat{x} =$
$\left[ \begin{array}{ccc} 36 \\ 112 \\ \end{array}\right] b$

You can use elimination and solve $4x_1 + 8x_2 = 36$  and $8x_1 + 26x_2 = 112$ to find $\hat{x} =$
$\left[ \begin{array}{ccc} 1 \\ 4 \\ \end{array}\right]$

We can also use $A^TA\hat{x} = A^Tb$ and algebra to get $\hat{x} = A^Tb / A^TA = A^Tb (A^TA)^{-1}$, which won't work because the matrices don't align, however $\hat{x} = (A^TA)^{-1} A^Tb$ does give 
$\left[ \begin{array}{ccc} 1 \\ 4 \\ \end{array}\right]$ (see below). I'm not sure why.

Also, if we use the R function solve on $A^TA$ and $A^Tb$ we get $\hat{x} =$
$\left[ \begin{array}{ccc} 1 \\ 4 \\ \end{array}\right]$ (see below).

```{r xhat}
invata <- solve(ata)
xhat <- invata %*% atb
xhat
solve(ata, atb)
```


#### c) What is the squared error of this solution?

$E = ||Ax - b||^2 = \left[ \begin{array}{ccc} 1 \\ -3 \\ 5 \\ -3 \\ \end{array}\right]$ for $x = \left[ \begin{array}{ccc} 1 \\ 4 \\ \end{array}\right]$

#### d) Instead of b = [0; 8; 8; 20], start with p = [1; 5; 13; 17] and find the exact solution

The system Ax = p is given below:

$\left[ \begin{array}{ccc} 1 & 0 \\ 1 & 1 \\ 1 & 3 \\ 1 & 4 \\ \end{array}\right]$
$\left[ \begin{array}{ccc} x_1 \\ x_2 \\ \end{array}\right] =$
$\left[ \begin{array}{ccc} 1 \\ 5 \\ 13 \\ 17 \\ \end{array}\right]$

which is solved by $x_1 = 1, x_2 =4$

```{r xact}
p <- matrix(c(1, 5, 13, 17), nrow=4)
atp <- t(A) %*% p
atp
solve(ata, atp)
```


#### e) Show that the error e = b - p = [-1, 3, -5, 3]

$b = \left[ \begin{array}{ccc} 0 \\ 8 \\ 8 \\ 20 \\ \end{array}\right]$

$p = \left[ \begin{array}{ccc} 1 \\ 5 \\ 13 \\ 17 \\ \end{array}\right]$

$b - p = \left[ \begin{array}{ccc} 0 - 1 \\ 8 - 5 \\ 8 - 13 \\ 20 - 17 \\ \end{array}\right] =$
$\left[ \begin{array}{ccc} -1 \\ 3 \\ -5 \\ 3 \\ \end{array}\right]$

```{r err}
e = b - p
e
```


#### f) Show that the error e is orthogonal to p and to each of the columns of A

Two vectors are orthogonal when their dot product is zero: $\mathbf{v} \cdot \mathbf{w} = 0$ or 
$\mathbf{v}^T \mathbf{w} = 0$

```{r serr}
t(e) %*% p
t(e) %*% A[,1]
t(e) %*% A[,2]
```


## Problem Set 2

### Write an R markdown script that takes in the auto-mpg data, extracts an A matrix from the first 4 columns and b vector from the fifth (mpg) column.


```{r mpg}
dfrm <- read.table("C:/Users/robgo/Documents/CUNY/2016-2-Fall-IS605-Comp-Math/Week5/auto-mpg.data")
A <- matrix(0, 392, 4)
A[,1] <- dfrm$V1
A[,2] <- dfrm$V2
A[,3] <- dfrm$V3
A[,4] <- dfrm$V4
b <- dfrm$V5
head(A)
tail(A)
head(b)
tail(b)
```


### calculate the fitting error between the predicted mpg of your model and the actual mpg

```{r fit}
ata <- t(A) %*% A
atb <- t(A) %*% b
solve(ata, atb)
```


